Segmentation-registration of diffusion MR images of the brain driven by active-contours

### 2015-02-17 Publications/2015-NeuroImage/03-introduction.tex
Update all the sections

% -*- root: 00-main.tex -*-
\section{Introduction}
\label{sec:introduction}

Nonlinear registration of images is a challenging task, ubiquitous
  in an endless number of image analysis applications.
The process aims to find a spatial mapping $U$ that aligns the information available
  in two different coordinate systems (generally called
  \emph{reference}, $R$, and \emph{moving}, $M$):
  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \citeauthor{sotiras_deformable_2013} \citep{sotiras_deformable_2013}
  illustrates the wide variety of existing registration methodologies.
They classify the algorithms by the three principal building blocks that can be identified
  among methods: the matching criteria or similarity function to evaluate the proximity of
  the solution, the deformation model that may implement theoretical properties and constraints
  of the mapping in the selected application, and the optimization method or search strategy.
The possibly most widespread matching criteria are those classified as \emph{iconic
  methods} in \citep{sotiras_deformable_2013}.
These algorithms require the definition of an appropriate function on the voxel-wise information
  content of both images.
A second group, called \emph{geometric methods}, measure distances between features derived from
  corresponding objects defined in both spaces, such as landmarks, shapes, surfaces, etc.
Finally, \emph{hybrid methods} combine properties of the iconic and geometric methods.
As reviewed in \citep{sotiras_deformable_2013}, an important research effort has been devoted to
  study application-oriented matching criteria.

In this work we propose a hybrid approach to nonlinear registration of anatomically correct
  and geometrically precise surfaces to low-resolution and distorted \gls*{dmri} of the brain.
Currently, we are witnessing an uprising demand for such algorithms in the challenging context
  of extracting and characterizing the structural connectivity networks of the
  human brain \citep{craddock_imaging_2013}.
Our registration method is demonstrated here as a solution to correct for the susceptibility
  distortions that \gls*{dmri} of the head typically present \cite{jezzard_correction_1995}.
A fundamental antecedent of this approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align \gls*{t1} and \gls*{dmri} images
  of the brain using a linear transformation model.
As in \citep{greve_accurate_2009}, we exploit the precise structures that can be extracted with
  widely-used software (i.e. \emph{FreeSurfer}~\citep{fischl_freesurfer_2012}) at high resolution,
  and then map this prior information to low resolution data, in which the scale of the imaged
  structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \cite{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving coordinate system.

>>>>

% -*- root: 00-main.tex -*-
\section*{Introduction}
\label{sec:introduction}

Nonlinear registration of images is a challenging task, ubiquitous
  in an endless number of image analysis applications.
The process aims to find a spatial mapping $U$ that aligns the information available
  in two different coordinate systems (generally called
  \emph{reference}, $R$, and \emph{moving}, $M$):
  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \cite{sotiras_deformable_2013}
  illustrates the wide variety of existing registration methodologies.
They classify the algorithms by the three principal building blocks that can be identified
  among methods: the matching criteria or similarity function to evaluate the proximity of
  the solution, the deformation model that may implement theoretical properties and constraints
  of the mapping in the selected application, and the optimization method or search strategy.
The possibly most widespread matching criteria are those classified as \emph{iconic
  methods} in \citep{sotiras_deformable_2013}.
These algorithms require the definition of an appropriate function on the voxel-wise information
  content of both images.
A second group, called \emph{geometric methods}, measures distances between features derived from
  corresponding objects defined in both spaces, such as landmarks, shapes, surfaces, etc.
Finally, \emph{hybrid methods} combine properties of the iconic and geometric methods.
As reviewed in \citep{sotiras_deformable_2013}, an important research effort has been devoted to
  study application-oriented matching criteria.

In this work we propose a hybrid approach to nonlinear registration of anatomically correct
  and precise surfaces to warped and low-resolution \gls*{dmri} of the brain.
Currently, we are witnessing an uprising demand for such algorithms in the challenging context
  of extracting and characterizing the structural connectivity networks of the
  human brain \citep{craddock_imaging_2013}.
Our registration method is demonstrated here as a solution to correct for the susceptibility
  distortions that \gls*{dmri} of the head typically present \citep{jezzard_correction_1995}.
\todo[inline]{talk about the distortion and cite SDCs}

A fundamental antecedent of the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align \gls*{t1} and \gls*{dmri} images
  of the brain using a linear transformation model.
Similarly, we exploit the precise structures that can be extracted with
  widely-used software (i.e. \emph{FreeSurfer}~\citep{fischl_freesurfer_2012}),
  and then map these surfaces into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \cite{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving coordinate system.

### 2015-02-17 Publications/2015-NeuroImage/04-methods.tex
Update all the sections

\section{Methods}
\label{sec:methods}
%
\subsection{Related work}
\label{sec:methods_background}
Registration-segmentation methods are typically built including a deformation
  model to guide the evolution of a segmentation algorithm.
Even though some Bayesian approaches have been proposed to jointly solve the problem
  \citep{wyatt_map_2003,gass_simultaneous_2014}, segmentation is usually performed using
  an evolving active contour \citep{chan_active_2001} that normally includes a shape prior
  term to the energy functional \citep{chen_using_2002,bresson_variational_2006,
  chan_level_2005,cremers_kernel_2006,gastaud_combining_2004}.
These methods generally have an inplicit description of the relative boundary locations of the objects
  to be delineated using level sets, and some even model the statistical deviations from original
  shape.

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000},
  and the first proper registration method was presented by \citeauthor{yezzi_variational_2001}
  \citep{yezzi_variational_2001}, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping
  (the so-called ``2 \glspl*{pde} approach'').
\citeauthor{vemuri_joint_2003} proposed the first atlas-based registration \citep{vemuri_joint_2003}
  based on the level-sets framework and using only one \gls*{pde}.
The principal application of these techniques was the image registration between timesteps in
  a time-series or between different images \citep{paragios_level_2003}.
\citeauthor{unal_coupled_2005} \citep{unal_coupled_2005} extended the idea of
  \citep{bertalmio_morphing_2000,yezzi_variational_2001} to nonlinear registration
  using a free deformation field model.
A summary of these second set of techniques is performed in \citep{droske_mumfordshah_2009},
  proposing two different approaches to applying the Mumford-Shah functiona
  \citep{mumford_optimal_1989} in simultaneous registration-segmentation, through the propagation
  of the deformation field from the contours to the whole image definition.
Finally, \citeauthor{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies \citep{gorthi_active_2011} based on the level-sets approximation.
Even though the historical evolution of joint segmentation and registration procedures is strongly
  linked to the level sets approach, our method conversely used the explicit representation of
  the evolving surfaces.

Finally, a derivation of the latter group is composed by atlas-based segmentation methods
  \citep{gorthi_segmentation_2009,gorthi_active_2011,pohl_unifying_2005,pohl_bayesian_2006,
  wang_joint_2006}, where the prior imposes consistent voxel-based classification of contiguous
  regions.
A comprehensive summary of the methodologies derived from the \acrlong{acwe} model, with special attention
  to joint registration-segmentation methods is found in \citep{gorthi_active_2011}.
is based in derivations of the \acrlong{acwe}
  framework \citep{chan_active_2001}, and related to \citep{guyader_combined_2011}.

\subsection{Cost-function derivation}
\label{sec:methods_map}

In a Bayesian framework, the mappings $U$ in \autoref{eq:transform} are
  evaluated based on their posterior probability given the observed data
  $R$.
Using the Bayes' rule, the posterior likelihood is computed as:
  \begin{equation}
  P(U \mid R,\Gamma_{l,m}) = \frac{P(R \mid U,\Gamma_{l,m})\, P(U)}{P(R)},
  \label{eq:bayes_rule}
  \end{equation}
  where $P(R \mid U,\Gamma_{l,m})$ is the data-likelihood, and
  $\Gamma_{l,m}$ are a set of surfaces corresponding to the interfaces
  between piecewise smooth regions $\Omega_i$, such that
  $\Gamma_{l,m} = \partial \Omega_l \cap \partial \Omega_m$ is the
  contour between tissues $l$ and $m$.
Then, the \gls*{map} criterium \citep{leemput_automated_1999} is used
  to find the mapping estimate $\hat{U}$ that aligns $\Gamma$ into $R$:
  \begin{align}
  \hat{U} &= \underset{U}{\argmax} \left\{ P(U \mid R,\Gamma_{l,m}) \right\} \notag\\
   &= \underset{U}{\argmax} \left\{ P(R \mid U,\Gamma_{l,m})\, P(U) \right\}.
  \label{eq:map_u}
  \end{align}

First, we assume independence between pixels, and thus break down the
  global data likelihood into a product of pixel-wise conditional probabilities:
  \begin{equation}
  P(R \mid U,\Omega) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
  where $\vec{f}' = R(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ (\autoref{eq:transform}) of the multispectral target
  volume.
For convenience, and because it has been shown to be appropriate approximations
  \citep{cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions:
  \begin{equation}
  P\left( \vec{f}' \mid U,\Omega_l \right) = \mathcal{N} \left( \vec{f}' \mid \Theta_l \right),
  \label{eq:likelihood}
  \end{equation}
 	where $\Theta_l = \lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$ are the
 	descriptors of region $\Omega_l$.
We then insert \autoref{eq:likelihood} into \autoref{eq:bayes_rule} to obtain the full
  formulation of the \emph{data term} of our registration model:
 	\begin{align}
  P( R \mid U, \Omega ) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) \notag\\
  &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mathcal{E}_l(\vec{r}) = \mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.

@@

\subsection{Numerical Implementation}
\label{sec:numerical_implementation}

@@

\subsubsection{Deformation model}
\label{sec:deformation_model}
Let us denote $\{\vec{v}_i\}_{i=1 \ldots N_c}$ the vertices of one or several prior
  surface(s).
In our application, these surfaces are triangularized meshes extracted using \emph{FreeSurfer}
  \citep{fischl_freesurfer_2012}.
The transformation from the structural space into the coordinate space of the
  target image is achieved through a dense deformation field $u(\vec{r})$, such that:
  \begin{equation}
  \vec{v}_i' = U\{\vec{v}_i\} = \vec{v}_i + u(\vec{v}_i),
  \label{eq:nodes_tfm}
  \end{equation}
  where $U$ is defined in \eqref{eq:transform}.
Since the nodes of the anatomical surfaces likely lay off-grid, it is required to
  derive $u(\vec{r})$ from a discrete set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved through a set of associated basis functions $\psi_k$:
  \begin{equation}
  u(\vec{r}) = \sum_k \psi_k(\vec{r}) \vec{u}_k.
  \label{eq:intp_kernel}
  \end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
  of degree 3 ($B_3$).
Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
  $\psi$ by the actual kernel function, the transformation writes:
  \begin{equation}
    \vec{v}_i' = \vec{v}_i + \sum_k \left[ \vec{u}_k \, \underset{d}{\prod}
      B_3( (\vec{v}_i - \vec{r}_k) \cdot \hat{\mathbf{e}}_d ) \right],
  \label{eq:transformation}
  \end{equation}
  with $\hat{\mathbf{e}}_d$ being the unitary vector along axis $d$.

\subsubsection{Optimization}
\label{sec:gradient_descent}
To find the minimum of the energy functional \eqref{eq:energy},
  we propose a gradient-descent approach with respect to the underlying
  deformation field through the following \gls*{pde}:
  \begin{equation}
  \frac{\partial u(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
  with $t$ being an artificial time parameter of the contour
  evolution, and $\vec{u}_k$ the parameters supporting the estimate
  $\hat{U}$ of the transformation at the current time point.
Now, we introduce \eqref{eq:energy} in \eqref{eq:general_gradient_descent}:
  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \int_{\Omega} \underset{l}{\sum} \mdist{f'}{l} \,d\vec{r} \notag\\
  &+ \int_{\Omega} [ \boldsymbol{\alpha} \cdot u(\vec{r})^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \cdot u(\vec{r}))^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{align}

@@

\subsection{Experiments and evaluation}
\label{sec:experiments_evaluation}
%
The experiments follow a five-steps design:
\begin{enumerate}
	\item The prior surfaces are extracted from an undistorted and high-resolution dataset
	  (e.g. \autoref{fig:phantom}).
	\item A plausible, synthetic distortion $U^{-1}_{true}$ is generated.
	\item The low-resolution (target) images are warped using the synthetic distortion.
	\item Our method is run, obtaining a $\hat{U}_{tst}$.
	  A cross-comparison methodology is also applied, to obtain a competing $\hat{U}_{cc}$.
	\item Results are visually and quantitatively evaluated.
\end{enumerate}

To quantitatively evaluate the registration results we computed the \gls*{swindex} as the one-to-one
  distance between corresponding vertices of surfaces, weighted by their respective
  Voronoi area $w_i$:
  \begin{equation}
  sWI = \frac{1}{P} \sum\limits_p^P \frac{1}{A_p} \sum\limits_i^{N_p} w_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \end{equation}
  where $\vec{v}_i$ are the locations of the $N_p$ vertices in each $p \in \{1, \dots P\}$
    prior, $A_p$ the total area of surface $p$, and $\hat{\vec{v}}_i$ is the location
    recovered corresponding to the vertex $\vec{v}_i$.

In the application on real data, three experts blindly rated correction results obtained
  with the proposed method and the cross-comparison reference.
The full report of visual results is provided within the Supplemental Materials.

  % \item \gls{wi}, L2-distance between the theoretical and the recovered
  %   deformation field.
  %   \begin{equation}
  %   WI = \frac{1}{N} \sum\limits_i^N \| \mathbf{d}_i - \hat{\mathbf{d}}_i \|
  %   \end{equation}
  %   where $\mathbf{d}_i$ is the theoretical displacement vector at position $i$
  %   and $\hat{\mathbf{d}}_i$ is the recovered one at the same index position.
  % \item Parcellation agreement, the Dice overlap index of the \glspl*{roi}
  %   contained by the parcellation.
  % \end{itemize}

\subsection{Image Data and preprocessing}
\label{sec:datasets}

\subsubsection{Simulated digital phantom} %
\label{sec:digital_phantoms}
%
We first present the proof of the concept on a simplistic digital phantom
  (\autoref{fig:phantom}).
Using \cite{caruyer_phantomas_2014}, we simulated 2.0mm isotropic \gls*{t1} and
  \gls*{t2} images at several rician noise levels, with parameters TE/TR=10/1500ms
  and TE/TR=90/5000ms respectively.
The reference surfaces were extracted using the marching-cubes algorithm
  shipped with \emph{FreeSurfer} \citep{fischl_freesurfer_2012}
  on high-resolution (1.0mm isotropic) and binary images corresponding to
  the tissue fraction maps used in simulations.

@@

\subsubsection{Real \gls{mri} datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm to real \gls*{dmri} data of human brains,
  we downloaded 10 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for preprocessing issues.
The datasets comprehend \gls*{t1}, \gls*{t2} and multi-shell \gls*{dmri} images
  already corrected for artifacts, brain-extracted and spatially normalized,
  along with the most prominent results of the standard processing with
  \emph{FreeSurfer}: \emph{aparc} segmentations, and the surfaces corresponding to
  the cortical \gls*{gm} exterior and the \gls*{wm} interface with
  cortical \gls*{gm}.

>>>>

\section*{Methods}
\label{sec:methods}
%
\subsection*{Related work}
\label{sec:methods_background}
Registration-segmentation methods are generally built including a deformation model to
  support the evolution of a segmentation algorithm that drives the mapping process.
Thus, solutions usually select first a segmentation strategy to derive the cost function and
  regularization of the problem.
Generally, evolving active contours is the most extended segmentation approach to be
  integrated within the registration process.
These methods either have an implicit description of the relative boundary locations of the
  objects to be delineated using level sets \citep{chen_using_2002,bresson_variational_2006,
  chan_level_2005}, or model the statistical deviations from original shape
  \citep{cremers_kernel_2006,gastaud_combining_2004}.
Some Bayesian approaches have been proposed as well \citep{wyatt_map_2003,gass_simultaneous_2014}.
In \autoref{sec:methods_map} we start from a probabilistic framework to finally propose
	a cost function defined on the location of explicit surfaces, showing the duality of the
	existing approaches.

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Soonly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping
  (the so-called ``2 \glspl*{pde} approach'').
\cite{vemuri_joint_2003} proposed the first atlas-based registration based on the level sets
  framework and using only one \gls*{pde}.
The principal application of these techniques was the image registration between timesteps in
  a time-series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} extended the idea of \citep{yezzi_variational_2001} to nonlinear
  registration using a free deformation field model.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on the level sets approximation.

\todo[inline]{Revise these references and include if necessary:
  \citep{pohl_unifying_2005,pohl_bayesian_2006,wang_joint_2006,guyader_combined_2011}}

\subsection*{Cost-function derivation}
\label{sec:methods_map}

In a Bayesian framework, the mappings $U$ in \autoref{eq:transform} are
  evaluated based on their posterior probability given the observed data
  $R$.
Using the Bayes' rule, the posterior likelihood is computed as:
  \begin{equation}
  P(U \mid R,\Gamma_{l,m}) = \frac{P(R \mid U,\Gamma_{l,m})\, P(U)}{P(R)},
  \label{eq:bayes_rule}
  \end{equation}
  where $P(R \mid U,\Gamma_{l,m})$ is the data-likelihood, and
  $\Gamma_{l,m}$ are a set of surfaces corresponding to the interfaces
  between piecewise smooth regions $\Omega_i$, such that
  $\Gamma_{l,m} = \partial \Omega_l \cap \partial \Omega_m$ is the
  contour between tissues $l$ and $m$.
Then, the \gls*{map} criterium is applied to find the mapping estimate $\hat{U}$
  that aligns $\Gamma$ into $R$:
  \begin{align}
  \hat{U} &= \underset{U}{\argmax} \left\{ P(U \mid R,\Gamma_{l,m}) \right\} \notag\\
   &= \underset{U}{\argmax} \left\{ P(R \mid U,\Gamma_{l,m})\, P(U) \right\}.
  \label{eq:map_u}
  \end{align}

First, we assume independence between pixels, and thus break down the
  global data likelihood into a product of pixel-wise conditional probabilities:
  \begin{equation}
  P(R \mid U,\Omega) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
  where $\vec{f}' = R(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ (\autoref{eq:transform}) of the multispectral target
  volume.
For convenience, and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions:
  \begin{equation}
  P\left( \vec{f}' \mid U,\Omega_l \right) = \mathcal{N} \left( \vec{f}' \mid \Theta_l \right),
  \label{eq:likelihood}
  \end{equation}
 	where $\Theta_l = \lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$ are the
 	descriptors of region $\Omega_l$.
We then insert \autoref{eq:likelihood} into \autoref{eq:bayes_rule} to obtain the full
  formulation of the \emph{data term} of our registration model:
 	\begin{align}
  P( R \mid U, \Omega ) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) \notag\\
  &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mathcal{E}_l(\vec{r}) = \mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.

@@

\subsection*{Numerical Implementation}
\label{sec:numerical_implementation}

@@

\paragraph*{Deformation model}
\label{sec:deformation_model}
Let us denote $\{\vec{v}_i\}_{i=1 \ldots N_c}$ the vertices of one or several prior
  surface(s).
In our application, these surfaces are triangularized meshes extracted using \emph{FreeSurfer}
  \citep{fischl_freesurfer_2012}.
The transformation from the structural space into the coordinate space of the
  target image is achieved through a dense deformation field $u(\vec{r})$, such that:
  \begin{equation}
  \vec{v}_i' = U\{\vec{v}_i\} = \vec{v}_i + u(\vec{v}_i),
  \label{eq:nodes_tfm}
  \end{equation}
  where $U$ is defined in \eqref{eq:transform}.
Since the nodes of the anatomical surfaces likely lay off-grid, it is required to
  derive $u(\vec{r})$ from a discrete set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved through a set of associated basis functions $\psi_k$:
  \begin{equation}
  u(\vec{r}) = \sum_k \psi_k(\vec{r}) \vec{u}_k.
  \label{eq:intp_kernel}
  \end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
  of degree 3 ($B_3$).
Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
  $\psi$ by the actual kernel function, the transformation writes:
  \begin{equation}
    \vec{v}_i' = \vec{v}_i + \sum_k \left[ \vec{u}_k \, \underset{d}{\prod}
      B_3( (\vec{v}_i - \vec{r}_k) \cdot \hat{\mathbf{e}}_d ) \right],
  \label{eq:transformation}
  \end{equation}
  with $\hat{\mathbf{e}}_d$ being the unitary vector along axis $d$.

\paragraph*{Optimization}
\label{sec:gradient_descent}
To find the minimum of the energy functional \eqref{eq:energy},
  we propose a gradient-descent approach with respect to the underlying
  deformation field through the following \gls*{pde}:
  \begin{equation}
  \frac{\partial u(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
  with $t$ being an artificial time parameter of the contour
  evolution, and $\vec{u}_k$ the parameters supporting the estimate
  $\hat{U}$ of the transformation at the current time point.
Now, we introduce \eqref{eq:energy} in \eqref{eq:general_gradient_descent}:
  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \int_{\Omega} \underset{l}{\sum} \mdist{f'}{l} \,d\vec{r} \notag\\
  &+ \int_{\Omega} [ \boldsymbol{\alpha} \cdot u(\vec{r})^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \cdot u(\vec{r}))^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{align}

@@

\subsection*{Experiments and evaluation}
\label{sec:experiments_evaluation}
%
In order to demonstrate the applicability of the registration method, we first
  subjected it to a battery of random tests over synthetic data and then applied
  it on real data.
\autoref{fig:evworkflows} presents the workflow implementing the evaluation instruments.
Besides a visual assessment of the results, we report quantitative evaluations using
  two metrics.
In the case of the phantoms, as we warped data along the three available dimensions,
  we computed the Haussdorf distance, integrating the ``point-to-cell'' method
	implemented by \cite{commandeur_vtk_2011}.
Conversely, the susceptibility-derived distortions only happen along the phase
  encoding axis of the image.
Therefore, a \gls*{swindex} can be computed as the one-to-one distance between corresponding
  vertices of surfaces, weighted by their respective Voronoi area $w_i$:
  \begin{equation}
  sWI = \frac{1}{P} \sum\limits_p^P \frac{1}{A_p} \sum\limits_i^{N_p} w_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|
  \end{equation}
  where $\vec{v}_i$ are the locations of the $N_p$ vertices in each $p \in \{1, \dots P\}$
  prior, $A_p$ the total area of surface $p$, and $\hat{\vec{v}}_i$ is the location
  recovered corresponding to the vertex $\vec{v}_i$.

\begin{figure}
\caption{1) The prior surfaces are extracted from an undistorted and high-resolution dataset
	  (e.g. \autoref{fig:phantom}). 2) A plausible, synthetic distortion $U^{-1}_{true}$ is generated.
	3) The low-resolution (target) images are warped using the synthetic distortion.
	4) Our method is run, obtaining a $\hat{U}_{tst}$.
	  A cross-comparison methodology is also applied, to obtain a competing $\hat{U}_{cc}$.
	5) Results are visually and quantitatively evaluated.}\label{fig:evworkflows}
\end{figure}

\subsection*{Image Data and preprocessing}
\label{sec:datasets}

\paragraph*{Simulated phantoms}%
\label{sec:digital_phantoms}
We proved the concept on simplistic digital phantoms which we will name after their
  appearance as ``box'', ``ball'', ``L-shape'', and ``gyrus'' (\autoref{fig:phantom}).
Using \emph{phantomas} \citep{caruyer_phantomas_2014}, we simulated \gls*{t1} and
  \gls*{t2} images (TE/TR=10/1500ms and TE/TR=90/5000ms
  respectively) corresponding to each shape.
Simulations were conducted with rician noise for a \gls*{snr} of 300.0, and
  two image resolutions, $1.0mm$ and $2.0mm$, both isotropic.
The reference surfaces were extracted using the marching-cubes algorithm
  shipped with \emph{FreeSurfer} \citep{fischl_freesurfer_2012}
  at the higher resolution (1.0mm isotropic)..

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 15 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend \gls*{t1}, \gls*{t2} and multi-shell \gls*{dmri} images
  already corrected for artifacts, brain-extracted and spatially normalized,
  along with the most prominent results of the standard processing with
  \emph{FreeSurfer}: \emph{aparc} segmentations, and the surfaces corresponding to
  the cortical \gls*{gm} exterior and the \gls*{wm} interface with
  cortical \gls*{gm}.
